---
title: "Day 1 — Spatial Data Retrieval & Analysis"
subtitle: "From APIs to clean, analysis-ready datasets in R"
author: "Your Name"
format: revealjs
bibliography: references.bib
---

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(sf)
library(terra)
library(rgbif)
library(CoordinateCleaner)
library(rnaturalearth)
library(tidyverse)
library(ggspatial)

theme_set(theme_minimal(base_size = 13))
```

## Spatial Data: A Primer {background-color="#003366" .white-text}

## What is Spatial Data?

> Data that explicitly represents **location** on Earth's surface,
> using coordinates within a defined **reference system**.

:::: {.columns}

::: {.column width="50%"}
### Vector (`sf`)

- Points → occurrences
- Lines → rivers, roads
- Polygons → protected areas
:::

::: {.column width="50%"}
### Raster (`terra`)

- Continuous gridded surfaces
- Elevation, temperature, NDVI
- Each cell = one value
:::

::::

::: aside
Comprehensive treatment in @lovelace2019geocomputation and @pebesma2023spatial.
:::

## Coordinate Reference Systems

The mathematical framework that links coordinates to locations on Earth.

$$
\text{CRS} = \{\,\text{datum},\ \text{projection},\ \text{units}\,\}
$$

```{r}
#| eval: false

# Always declare CRS explicitly with EPSG codes
st_crs(4326)   # WGS 84 — geographic, degrees
st_crs(3857)   # Web Mercator — projected, metres
st_crs(2100)   # GGRS87 / Greek Grid — projected, metres
```

::: {.callout-important}
## Always check the CRS

Mismatched CRS is the **#1 source of errors** in spatial analysis.
:::

## The `sf` Workflow

```{r}
#| eval: false

# Read country boundary
greece <- ne_countries(country = "Greece",
                       scale = "medium",
                       returnclass = "sf")

# Inspect
greece |> st_geometry() |> plot()

# Reproject
greece_2100 <- st_transform(greece, 2100)

# Compute area
greece_2100 |> st_area() |> units::set_units("km^2")
```

## The `terra` Workflow

```{r}
#| eval: false

library(terra)
library(geodata)

# Download WorldClim bioclimatic variables (10 arc-min)
bioclim <- worldclim_global(var = "bio", res = 10, path = "data/raw")

# Crop to Greece extent
greece_ext <- ext(19, 30, 34, 42)
bio_gr <- crop(bioclim, greece_ext)

# Plot annual mean temperature (BIO1)
plot(bio_gr[[1]], main = "Annual Mean Temperature (°C)")
```

## Retrieving GBIF Occurrences {background-color="#003366" .white-text}

## `rgbif`: Programmatic Access to GBIF

```{r}
#| eval: false

library(rgbif)

# 1. Resolve a taxon key
key <- name_backbone(name = "Quercus ilex")$usageKey

# 2. Search occurrences in Greece with valid coordinates
occ <- occ_search(
  taxonKey           = key,
  country            = "GR",
  hasCoordinate      = TRUE,
  hasGeospatialIssue = FALSE,
  limit              = 5000
)

# 3. Extract and convert to sf
pts <- occ$data |>
  select(species, decimalLongitude, decimalLatitude,
         year, basisOfRecord, datasetKey) |>
  drop_na(decimalLongitude, decimalLatitude) |>
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"),
           crs = 4326)
```

## Visualising Raw Occurrences

```{r}
#| eval: false

ggplot() +
  annotation_map_tile(type = "cartolight", zoom = 6) +
  geom_sf(data = greece, fill = NA, color = "grey30") +
  geom_sf(data = pts, color = "#4A7C59", alpha = 0.5, size = 1) +
  labs(title    = expression(italic("Quercus ilex") ~ "occurrences in Greece"),
       subtitle = "Source: GBIF.org",
       caption  = "CRS: EPSG:4326") +
  theme(plot.title = element_text(face = "bold"))
```

## Data Cleaning Pipeline

```{mermaid}
flowchart LR
  A[Raw GBIF<br/>records] --> B[Remove<br/>duplicates]
  B --> C[Flag<br/>centroids & 0,0]
  C --> D[Spatial<br/>thinning]
  D --> E[Validate<br/>taxonomy]
  E --> F[(Analysis-ready<br/>dataset)]

  style A fill:#B22234,color:white
  style F fill:#4A7C59,color:white
```

## Cleaning with `CoordinateCleaner`

```{r}
#| eval: false

library(CoordinateCleaner)

flags <- clean_coordinates(
  x       = as.data.frame(occ$data),
  lon     = "decimalLongitude",
  lat     = "decimalLatitude",
  species = "species",
  tests   = c("capitals", "centroids", "equal",
              "gbif", "institutions", "zeros", "seas")
)

# Keep only records that pass all tests
clean <- occ$data[flags$.summary, ]
```

::: {.callout-tip}
`clean_coordinates()` returns a logical matrix flagging each test
individually — inspect failures **before** discarding!
:::

## Spatial Thinning to Reduce Sampling Bias

```{r}
#| eval: false

# Thin to one record per ~5 km grid cell
thin_grid <- function(pts, cell_km = 5) {
  pts |>
    st_transform(2100) |>
    mutate(cell = paste(
      floor(st_coordinates(geometry)[, 1] / (cell_km * 1000)),
      floor(st_coordinates(geometry)[, 2] / (cell_km * 1000)))) |>
    group_by(cell) |>
    slice(1) |>
    ungroup() |>
    st_transform(4326)
}

pts_thin <- thin_grid(pts, cell_km = 5)
```

## Quality Control: Before vs After

```{r}
#| eval: false

library(patchwork)

p1 <- ggplot(greece) +
  geom_sf(fill = "grey95") +
  geom_sf(data = pts, color = "#B22234", alpha = 0.4, size = 0.8) +
  labs(title = paste("Raw —", nrow(pts), "records")) +
  theme_void()

p2 <- ggplot(greece) +
  geom_sf(fill = "grey95") +
  geom_sf(data = pts_thin, color = "#4A7C59", alpha = 0.6, size = 0.8) +
  labs(title = paste("Thinned —", nrow(pts_thin), "records")) +
  theme_void()

p1 + p2
```

## Hands-on Exercise 1 {background-color="#003366" .white-text}

### 🛠️ Your turn (45 min)

1. Pick a Mediterranean species (plant or animal)
2. Retrieve GBIF occurrences with `rgbif::occ_search()`
3. Apply `CoordinateCleaner::clean_coordinates()`
4. Map raw vs cleaned distribution side-by-side
5. Briefly discuss flagged records with your neighbour

⏱️ We reconvene at **14:15**

## Take-aways from Day 1

- The R spatial stack (`sf` + `terra`) covers the full data pipeline
- Always verify and harmonise **CRS** before any spatial operation
- GBIF data are powerful but **noisy** — clean systematically
- Document every step → **reproducibility is non-negotiable**

::: {.callout-note}
Tomorrow we move from *where* species occur to *how diverse* assemblages are.
:::

## References

::: {#refs}
:::