---
title: "Day 2 — Biodiversity & Biogeographical Patterns"
subtitle: "Quantifying diversity and modelling distributions in R"
author: "Your Name"
format: revealjs
bibliography: references.bib
---

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(sf)
library(terra)
library(vegan)
library(betapart)
library(iNEXT)
library(predicts)
library(blockCV)
library(usdm)
library(tidyverse)
library(patchwork)
library(viridis)

theme_set(theme_minimal(base_size = 13))
```

## Quantifying Biodiversity {background-color="#003366" .white-text}

## The α, β, γ Framework

Following @whittaker1972evolution:

$$
\gamma = \alpha \times \beta
$$

| Component   | Scale                | R function              |
|-------------|----------------------|-------------------------|
| α (alpha)   | Local richness       | `vegan::specnumber()`   |
| β (beta)    | Turnover among sites | `betapart::beta.pair()` |
| γ (gamma)   | Regional richness    | `vegan::specpool()`     |

## Hill Numbers — A Unified Framework

$$
{}^{q}\!D = \left( \sum_{i=1}^{S} p_i^{\,q} \right)^{1/(1-q)}
$$

- \(q = 0\): species richness
- \(q = 1\): exponential of Shannon entropy
- \(q = 2\): inverse Simpson concentration

```{r}
#| eval: false

library(iNEXT)
out <- iNEXT(community_data, q = c(0, 1, 2),
             datatype = "abundance", endpoint = 500)
ggiNEXT(out, type = 1) + scale_color_viridis_d(option = "D")
```

## α-Diversity Example

```{r}
#| eval: false

library(vegan)
data(BCI)   # tropical tree community

div <- tibble(
  site     = rownames(BCI),
  S        = specnumber(BCI),
  H        = diversity(BCI, index = "shannon"),
  Simpson  = diversity(BCI, index = "simpson"),
  Pielou_J = H / log(S)
)

div |>
  pivot_longer(-site) |>
  ggplot(aes(value)) +
  geom_histogram(fill = "#4A7C59", bins = 20) +
  facet_wrap(~ name, scales = "free")
```

## β-Diversity: Turnover vs Nestedness

@baselga2010partitioning decomposed Sørensen β into two components:

$$
\beta_{\text{SOR}} = \beta_{\text{SIM}} + \beta_{\text{SNE}}
$$

```{r}
#| eval: false

library(betapart)

# Convert abundance to presence/absence
pa <- ifelse(BCI > 0, 1, 0)

# Pairwise multi-component beta
bp <- beta.pair(pa, index.family = "sorensen")

# Mean values
sapply(bp, mean)
#>  beta.sim  beta.sne  beta.sor
#>      0.41      0.06      0.47
```

## Visualising β-Diversity

```{r}
#| eval: false

library(ape)

# UPGMA on turnover
hc <- hclust(bp$beta.sim, method = "average")

# NMDS ordination
nmds <- metaMDS(bp$beta.sor, k = 2, trymax = 50)

plot(nmds, type = "t", main = "NMDS — Sørensen dissimilarity")
```

## Species Distribution Modelling {background-color="#003366" .white-text}

## SDM Workflow

```{mermaid}
flowchart LR
  A[Cleaned<br/>occurrences] --> C[Predictor<br/>selection]
  B[Environmental<br/>layers] --> C
  C --> D[Pseudo-absences<br/>or background]
  D --> E[Model fitting]
  E --> F[Spatial<br/>cross-validation]
  F --> G[Prediction &<br/>uncertainty]

  style G fill:#4A7C59,color:white
```

## Predictor Selection — Multicollinearity

```{r}
#| eval: false

library(usdm)

# Variance Inflation Factor stepwise selection
v <- vifstep(bio_gr, th = 5)
v
#> 4 variables from 19 input variables have collinearity problem:
#>   bio6 bio11 bio16 bio17

# Keep only retained predictors
preds <- bio_gr[[v@results$Variables]]
```

::: {.callout-tip}
A VIF threshold of **5–10** is common; lower is stricter.
:::

## Background / Pseudo-absences

```{r}
#| eval: false

library(predicts)

# 10,000 random background points within Greece
set.seed(42)
bg <- spatSample(preds, size = 10000,
                 method = "random", na.rm = TRUE,
                 as.points = TRUE, values = FALSE)

# Extract environmental values at presence + background
pres_env <- extract(preds, vect(pts_thin))
bg_env   <- extract(preds, bg)

train <- bind_rows(
  mutate(pres_env, occ = 1),
  mutate(as_tibble(bg_env), occ = 0)
) |> drop_na()
```

## Fitting a MaxEnt-style Model

```{r}
#| eval: false

library(predicts)

# MaxEnt via the `predicts` package (successor of `dismo`)
mod <- MaxEnt(
  x = preds,
  p = st_coordinates(pts_thin),
  removeDuplicates = TRUE
)

# Predict habitat suitability
suit <- predict(preds, mod, type = "cloglog")
plot(suit, col = viridis(100), main = "Habitat suitability")
```

## Spatial Block Cross-Validation

```{r}
#| eval: false

library(blockCV)

sb <- cv_spatial(
  x         = pts_thin,
  column    = "occ",
  k         = 5,
  size      = 50000,   # 50 km blocks
  selection = "random",
  iteration = 100
)

cv_plot(cv = sb, x = pts_thin)
```

::: {.callout-important}
Random CV inflates SDM performance by exploiting spatial autocorrelation.
**Always use spatial blocking** for honest evaluation.
:::

## Model Evaluation

```{r}
#| eval: false

library(predicts)

eval_metrics <- pa_evaluate(
  p = predict(mod, pres_env_test),
  a = predict(mod, bg_env_test)
)

eval_metrics
#>   AUC    : 0.87
#>   COR    : 0.42
#>   max TSS: 0.61
```

## Biogeographical Regionalisation {background-color="#003366" .white-text}

## From β-Diversity to Bioregions

The workflow:

1. Build a site × species matrix on a regular grid
2. Compute pairwise β-diversity (`betapart`)
3. Apply hierarchical clustering (`hclust`)
4. Cut the dendrogram → bioregions
5. Map back to space

::: aside
See @kreft2010framework and @daru2020endemism for ecoregion delineation methods.
:::

## Gridded β-Diversity in R

```{r}
#| eval: false

# Build a 0.5° grid
grid <- st_make_grid(greece, cellsize = 0.5, square = TRUE) |>
        st_as_sf() |>
        mutate(cell_id = row_number())

# Join occurrences to grid
sp_grid <- st_join(all_species_pts, grid)

# Site × species matrix
mat <- sp_grid |>
  st_drop_geometry() |>
  count(cell_id, species) |>
  pivot_wider(names_from = species, values_from = n,
              values_fill = 0) |>
  column_to_rownames("cell_id")

# Presence/absence + Simpson dissimilarity
pa  <- ifelse(mat > 0, 1, 0)
dis <- beta.pair(pa)$beta.sim
hc  <- hclust(dis, method = "average")
clusters <- cutree(hc, k = 5)
```

## Mapping Bioregions

```{r}
#| eval: false

grid_clust <- grid |>
  filter(cell_id %in% rownames(mat)) |>
  mutate(region = factor(clusters))

ggplot() +
  geom_sf(data = greece, fill = "grey95") +
  geom_sf(data = grid_clust, aes(fill = region),
          color = NA, alpha = 0.85) +
  scale_fill_viridis_d(option = "D", name = "Bioregion") +
  labs(title    = "Biogeographical regionalisation of Greece",
       subtitle = "Based on Simpson β-diversity (turnover)",
       caption  = "Data: GBIF · CRS: EPSG:4326") +
  theme_void()
```

## Case Study: Mediterranean Flora {background-color="#003366" .white-text}

## Putting It All Together

:::: {.columns}

::: {.column width="50%"}
### Question

How is plant β-diversity structured across the Mediterranean basin?

### Data

- 25 woody species (GBIF)
- WorldClim BIO1, BIO12, BIO15
- SoilGrids texture
:::

::: {.column width="50%"}
### Methods

- `rgbif` → retrieval
- `CoordinateCleaner` → QC
- `betapart` → turnover
- `predicts` → SDM
- `blockCV` → validation
:::

::::

## Hands-on Exercise 2 {background-color="#003366" .white-text}

### 🛠️ Your turn (90 min)

1. Choose 3 species from yesterday's exercise
2. Stack their cleaned occurrences
3. Build a site × species matrix on a 0.25° grid
4. Compute pairwise β with `betapart::beta.pair()`
5. Cluster and map 2–4 bioregions
6. Discuss: do they align with known phytogeographical zones?

⏱️ Reconvene at **15:00** for synthesis

## Synthesis & Take-aways

- R offers a complete, reproducible pipeline for spatial ecology
- Choose biodiversity metrics that **match your question**, not vice-versa
- SDMs are powerful but require **honest evaluation** (spatial CV)
- Biogeographical patterns emerge from β-diversity **structure**
- Document, version, share → **open science is good science**

## Further Reading

- @lovelace2019geocomputation — Geocomputation with R
- @pebesma2023spatial — Spatial Data Science with R
- @baselga2010partitioning — β-diversity partitioning
- @valavi2021modelling — modern SDM best practices
- @whittaker1972evolution — α/β/γ framework

::: {.callout-tip}
## Repository
Slides, code, and data: [github.com/kkougiou/seminar-patras](https://github.com/kkougiou/seminar-patras)
:::

## Thank You! {background-color="#003366" .white-text}

### Ευχαριστώ πολύ!

- 📧 kkougiou@aua.gr
- 🐙 github.com/kkougiou
- 🆔 ORCID: 0000-0002-6938-3025
- 🌐 kkougiou.github.io

*Questions, collaborations, follow-ups — always welcome.*

## References

::: {#refs}
:::